Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-07-13 text/sections/01-introduction.tex
re-writing

A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via funding shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
that high school students exposed to similar public school funding shocks are less likely to attend college \citep{jackson2021school}, and that these shocks induce public universities to rely more on tuition as a source of revenue \citep{bound2019public}.
\cite{NBERw27885} use the same methodology to show that increases in state funding lead to degree lower completion time and later-life debt for students.\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower completion rates at public universities.
}

Over the last half century, there have been drastic changes to the heirarchy of US higher education, where the top universities have become more selective for student applicants, and the average university less selective \citep{hoxby2009changing}.
Similarly, research faculty at US universities are increasingly concentrated among the graduates of highly selective PhD programmes \citep{wapman2022quantifying}.
Faculty composition is an interesting outcome on its own, but its relationship with higher education finances has so far not been studied.
The closest previous examples study  how university endowments react to negative endowment shocks \citep{brown2014endowment} and the 2008 recession \citep{turner2014impact}, and multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
%Lastly, instruction costs within individual universities vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.
The number of professors per student may be particularly important for quality of instruction and thus educational outcomes;
small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Though, the economics literature is not decided on whether contingent lecturers lead to better or worse outcomes in higher education.
Lecturers are more effective teachers than research professors at a highly selective university \citep{bettinger2010does,figlio2015tenure}, but students taught by part-time lecturers have worse student outcomes relative to full-time lecturers or faculty \citep{zhu2021limited,ehrenberg2005tenured}.

>>>>

A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via funding shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
that high school students exposed to similar public school funding shocks are less likely to attend college \citep{jackson2021school}, and that these shocks induce public universities to rely more on tuition as a source of revenue \citep{bound2019public}.
\textbf{Faculty outcomes surveyed by \cite{ehrenberg2003studying}.}
\cite{NBERw27885} use the same methodology to show that increases in state funding lead to degree lower completion time and later-life debt for students.\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower completion rates at public universities.
}

Over the last half century, there have been drastic changes to the heirarchy of US higher education, where the top universities have become more selective for student applicants, and the average university less selective \citep{hoxby2009changing}.
Similarly, research faculty at US universities are increasingly concentrated among the graduates of highly selective PhD programmes \citep{wapman2022quantifying}.
Faculty composition is an interesting outcome on its own, but its relationship with higher education finances has so far not been studied.
The closest previous examples study  how university endowments react to negative endowment shocks \citep{brown2014endowment} and the 2008 recession \citep{turner2014impact}, and multiple theoretically model university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
%Lastly, instruction costs within individual universities vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.
The number of professors per student may be particularly important for quality of instruction and thus educational outcomes;
small class sizes in secondary schools lead to increases in test scores \citep{angrist1999using}, and UK university students perform worse academically in particularly large classes \citep{bandiera2010heterogeneous}.
Though, the economics literature is not decided on whether contingent lecturers lead to better or worse outcomes in higher education.
\textbf{Beef up this part to focus more on negative effects on students.}
Lecturers are more effective teachers than research professors at a highly selective university \citep{bettinger2010does,figlio2015tenure}, but students taught by part-time lecturers have worse student outcomes relative to full-time lecturers or faculty \citep{zhu2021limited,ehrenberg2005tenured,jaeger2011examining}.

\textbf{Tenure research \citep{mcpherson1999tenure,mcpherson1983economics}}

### 2023-07-13 text/sections/04-empirical.tex
re-writing

\textbf{TO-DO:
    Add description the dollar-by-dollar, and log terms, of the first stage.
}
The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects.
We note columns (1) and (2) estimate that a funding shock (per student in the entire state) of 10\% is associated with 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

@@

\subsection{Percent Change and Rates of Substitution}
\label{sec:empirical-substituion}

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shock instrument is a strong predictor for the average university's level of state funding in either unit; a funding shock of \$-1,000 per student in the entire state leads to around \$-1,176 per student at the university, while a funding shock of -10\% leads to around -9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same to California and Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in how much the funding higher education in 1990, but most have experienced a decline on order of 30\%, so that the stagnation in state funding has been a percent change trend over this time period.
As such, I focus on regression specifications where the explanatory and outcome variables are $\log$ transformed (unless otherwise mentioned).

The estimand of a $\log-\log$ specification is a funding elasticity; a 1\% fall in state funding $X_{i,t}$ caused a $\beta$\% fall in outcome $Y_{i,t}$, as in \autoref{eqn:secondstage}.
If the estimates are negative, then conceptually we consider the outcome to be inferior for a university's budgeting decisions with respect to state funding: if they are funded with more state resources they spend less on outcome $Y_{i,t}$.
Conversely, outcome $Y_{i,t}$ is normal if the coefficient is positive.

The funding elasticities can be used to recover the rates of substitution between two outcomes.
For example, write $Y^1$ for the number of lecturers per student at the universities, and $Y^2$ for the number of full professors.
I use the above approaches to estimate the funding elasticities, where $\% \Delta$ denotes percent change.
\[ \beta_1 = \frac{\% \Delta Y^1}{\% \Delta X}
\text{, and }
\beta_2 = \frac{\% \Delta Y^2}{\% \Delta X} \]
As such, it is possible to recover the elasticity for substitution between lecturers and full professors by the universities via the respective funding elasticities.
\[ \frac{\% \Delta Y^1}{\% \Delta Y^2}
= \frac{\% \Delta Y^1 / \% \Delta X}{\% \Delta Y^2 / \% \Delta X}
= \frac{\beta_1}{\beta_2} \]
I present results for the rates of substitution between different levels of faculty by this approach, dividing the relevant coefficient estimates and presenting standard errors calculated by a non-parametric bootstrap.
% https://stackoverflow.com/questions/63777368/computing-the-standard-error-when-dividing-coefficients-of-different-regressions

\subsection{Effects in Years After the Funding Shock}

>>>>

The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects;
Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
\textbf{TO-DO:
    Add description the dollar-by-dollar, of the first stage.
}
We note columns (1) and (2) estimate that a funding shock (per student in the entire state) of 10\% is associated with 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

@@

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shock instrument is a strong predictor for the average university's level of state funding in either unit; a funding shock of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shock of -10\% leads to around -9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same to California and Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in how much the funding higher education in 1990, but most have experienced a decline on order of 30\%, so that the stagnation in state funding has been a percent change trend over this time period.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

\subsection{Effects in Years After the Funding Shock}
\label{sec:empirical-substituion}

### 2023-07-13 text/sections/05-results.tex
re-writing

Changes in state funding have clear effects on the composition of faculty public universities employ.
\autoref{tab:facultycount-shock-reg} presents OLS and IV estimates for the effecet of a change in state funding on the number of faculty at the university, individually by position (and for the total faculty), in both count of professors and in percent terms for faculty per student ratio.
An extra \$1,000 in state funding per student leads to the average university employing 6 fewer lecturers, with no discernible effect on the count of total professors (\autoref{tab:facultycount-shock-reg}, Panel A).
In percentage terms, a 10\% increase in state funding lead to a 4.4\% decrease in the number of lecturers per student, an increase of 14\% for both assistant and full professors, leading to a 0.6\% increase in the total number of faculty per student.
In the state of Illinois, using the count of professors represented in the IBHED databse, state funding is correlated with the number of faculty, but lacks precision to identify the exact effects given the 144 university-year observations in the panel for 2010--2021.

Incumbent professors are relatively unaffected by changes in the state funding for their university in the years following their hire.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 1\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion; there are little effects discernible from zero, so that faculty who are already at the university have little of the effects of funding cuts passed on to them.

Faculty composition is not only impacted in the same years as a state funding shock, but also for a number years following.
\autoref{fig:all-count-lp} shows local projection estimates, where the elasticity of total professor count per student in year $t+k$ with respect to state funding in year $t$ is around 0.1, for follow-on years $k = 0, \hdots, 3$.
These findings are in line with the observation that universities froze hiring for a couple of years in response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
For lecturers, we see a negative effect which lines up with two trends noted in \autoref{sec:trends}: funding for public universities (per student) is decreasing while relative usage of  lecturers increased rapidly.

Decreases in state funding (via funding shocks) affects not only the total counts of professors, but also the composition in the years after thee funding shock.
A 10\% decrease in state funding increases count of lecturers per student by 4\% for the first two years after the initial shock.
While count of assistant professors decreases 10 to 18.5\% three years later;
count of tenured professors decreases 9 to 6\% three years later.
Together, the total count of professors per student decreases around 1\% for three years after the initial funding shock, and the effect is not distinguishable from zero after this point.
\autoref{fig:lecturer-count-lp}, \ref{fig:assistant-count-lp}, \ref{fig:full-count-lp} represents these estimates, using the local projection method, as a series of impulse responses for each outcome of professor count.
These findings show that public universities increase (decrease) their count of tenure-track and tenured professors per student in years when revenues are more (less) plentiful, possibly by increasing hiring intensity or efforts to maintain incumbent professors.
In the same vein, when revenues are bountiful, count of tenure-track and tenured professors per student increases --- and the opposite in years of funding cuts.

\subsection{Rates of Substitution}
\label{sec:results-substituion}

Write here about the log results, including a figure for the university funding elasticities of faculty demand.
Describe here the relevance of the elasticities, and describe the estimates for elasticity of substitution.

>>>>

Changes in state funding have clear effects on the composition of faculty public universities employ.
\autoref{tab:facultycount-shock-reg} presents OLS and IV estimates for the effect of a change in state funding on the number of faculty at the university, individually by position (and for the total faculty), in both count of professors and in percent terms for faculty per student ratio.
An extra \$1,000 in state funding per student leads to the average university employing 6 fewer lecturers, with no discernible effect on the count of total professors (\autoref{tab:facultycount-shock-reg}, Panel A).
In percentage terms, a 10\% increase in state funding lead to a 4.4\% decrease in the number of lecturers per student, an increase of 14\% for both assistant and full professors, leading to a 0.6\% increase in the total number of faculty per student.
In the state of Illinois, using the count of professors represented in the IBHED databse, state funding is correlated with the number of faculty, but lacks precision to identify the exact effects given the 144 university-year observations in the panel for 2010--2021.

Faculty composition is not only impacted in the same years as a state funding shock, but also for a number years following.
\autoref{fig:count-lp} shows local projection estimates, as a series of impulse responses, where the estimate is the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$ for follow-on years $k = 0, \hdots, 10$.
A 10\% decrease in state funding increases count of lecturers per student by 4\% for the first two years after the initial shock.
While count of assistant professors decreases 10 to 18.5\% three years later;
count of tenured professors decreases 9 to 6\% three years later.
Together, the total count of professors per student decreases around 1\% for three years after the initial funding shock, and the effect is not distinguishable from zero after this point.

Incumbent professors are relatively unaffected by changes in the state funding for their university in the years following their hire.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 1\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero, so that faculty who are already at the university (incumbent faculty) seem to have no effects of state funding cuts passed on to them.

%These findings show that public universities increase (decrease) their count of tenure-track and tenured professors per student in years when revenues are more (less) plentiful, possibly by increasing hiring intensity or efforts to maintain incumbent professors.
%In the same vein, when funding increases, count of tenure-track and tenured professors per student increases --- and the opposite in years of funding cuts.
%These findings are in line with the observation that universities froze hiring for a couple of years in response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
%For lecturers, we see a negative effect which lines up with two trends noted in \autoref{sec:trends}: funding for public universities (per student) is decreasing while relative usage of lecturers increased substantially.

The composition of faculty changed at public universities 1990--2017 thanks to stagnating state funding, but there are little discernible effects on incumbent faculty.
This may imply that the rate that public universities hired faculty was impacted by state funding cuts.
I take recent data on the total faculty hired at US public universities over 2011--2021 (provided by \citealt{wapman2022quantifying}), to show that universities with lower state funding (per student) also hired fewer professors in the same time period --- shown in \autoref{fig:hiring-correlation}.
Similarly, the funding shock IV model shows that a $10$\% cut in per student state funding leads to $13$\% fewer faculty hires per student across the decade 2011--2021.\footnote{
    See \autoref{sec:appendix-hiring} for further details.
}
These results provide limited evidence that changes in faculty composition arose by disrupting hiring of new professors at public universities, corresponding to the results that incumbent faculty were not by meaningfully effected by the funding cuts.

### 2023-07-13 text/sections/06-discussion.tex
re-writing

The results show that the recent stagnation in state funding for higher education has affected the composition of faculty within public universities, away from tenure-track and tenured professors towards contingent lecturers.
At the same time, there are little discernible effects on individual professors hired in the years 2011-2021 at Illinois public universities, except for the salaries of first year lecturers.
The rate that faculty leave their university, and their rate of promotion between positions, are unaffected by the university revenues, so this leaves one primary channel to explain changes in faculty composition: hiring.

Between academic years the number of professors at a university changes: professors leave an institution (retiring, hired elsewhere, etc.) and other join (they are newly hired, end a sabbatical, etc.).
\autoref{sec:results-indiv} shows that most of these channels are unaffected, implying that falls in hiring for tenured-track and tenured faculty must explain most of the substitution towards lecturers away from tenured faculty.
Notably, faculty hiring is already know to be heavily impacted by budget shocks.
\cite{turner2014impact} documents the wide-spread practice of hiring freezes at universities in response to budget shocks around the 2008 recession.
Throughout the last decade, multiple such measures were taken by Illinois public universities in response to their deteriorating finances \citep{furlough2010}.
The University of Illinois\footnote{
    The University of Illinois includes three public university campuses: Urbana-Champaign, Chicago, Springfield.
}
did not receive the allocated state appropriations from the state of Illinois on time, so enacted cost-cutting measures to stay fiscally solvent.
In response, the university system placed a hold on all hiring for filling state-funded positions and promotions.

The data used so far measure faculty count, so cannot disentangle which of these channels is most affected by changes in state funding at a public university, yet \cite{wapman2022quantifying} provide a measure of the aggregated total faculty hires at universities over the time period 2011--2021.\footnote{
    \citep{wapman2022quantifying} made an aggregated version of their data open-source, making this sub-analysis possible.
}
\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{State Funding and Faculty Hired at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.8\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation.png}
\end{figure}

Universities with more state funding (for the entire decade) hired more professors (\autoref{fig:hiring-correlation.png}).
Similarly, the funding shock instrument shows that an increase in 10\% state funding leads to 13\% more faculty hires at a university across the decade 2011--2021 (see \autoref{tab:hiring-shock-reg}).\footnote{
    These results were produced by integrating the total count of faculty hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
    There were no observable differences in the hiring rate of male vs female faculty.
    See \autoref{sec:appendix-hiring} for further details.
}
It is clear that public universities' hiring is negatively impacted by falls in state funding, even using highly aggregated data for a total count of hiring across an entire decade.
While this analysis strengthens the case negative effects on hiring plays, it is crude for measuring exact magnitudes for the effect on faculty hiring.
I leave it to further research to delve deeper into the magnitudes that hiring plays as the primary causal mediator for the effects of falls in state funding on faculty composition at public universities.

>>>>

Universities received less state funding, relative to increasing enrolment, and reacted by increasing their reliance on lecturers relative to assistant and full professors in both the short and medium-run; these results imply that lecturers can be considered substitutes for professors in the tenure-system.\footnote{
    Additional results in Appendix \autoref{sec:appendix-substitution} formalises the marginal rate of substitution between lecturers and professors on tenure-track, showing how assistant and full professors were substituted for lecturers while there is no substitution between assistant and full professors.
}
A public university's state funding per student fell by an average of 36\%, and the number of lecturers per student increased by 113\% (fell by 17\% for assistant and 23\% full professors).
The estimates of elasticity for faculty count with respect to state funding imply that stagnating state funding explains about 40\% of the observed substitution towards lecturers and away from tenure-track and tenured faculty.

These effects are persistent: public higher education in the US was not exposed to one large instance of funding cuts, but systematic funding short-falls across the last three decades.
The long-term estimates (i.e., \autoref{fig:count-lp}) show that state funding cuts have effects on universities for multiple years after the initial funding cut, and these effects are compounded by the fact that public universities received persistent funding cuts for so many years.
The end result is large changes in faculty composition, the long-run trend in stagnating state funding leading to a long-run substitution towards lecturers at public universities.

At the same time, incumbent faculty in the state of Illinois are unaffected by the state funding cuts in terms of salary, promotion rate, and rate of leaving their faculty position.
These results imply that composition change arose by lower hiring of tenure-track faculty in the public university system.
The private university system did not grow in any corresponding amount, with the clear implication that there are fewer tenure-track openings at US universities.\footnote{
    While the number of PhD graduates has continued to rise \citep{aau2021survey}.
}
Securing academic employment on the tenure-track is more selective than ever, so that more PhD graduates will end up teaching as lecturers or leaving academia.
% Note: unclear effects on PhD students choosing to not enter acadmeia, or go towards private unis
% Would be possible to study with data on PhD students, and whether they are more/less likely to enter academia after budget cuts to their PhD institution.

The results here are squarely in line with theories explaining the increased competition and inequality between US universities.
\cite{urquiola2020markets} interprets the success of research in US higher education as a result of free market policies, where increased competition in a free market in the early 1900s led to improvements in university research among US private universities.
These changes have not just affected research, they have affected education, too.
The rate of selectivity among US universities has become increasingly polarised, where top (and mostly private) universities with the most resources have become more selective at the undergraduate level \citep{hoxby2009changing}, and increasingly dominate academia at the graduate level \citep{wapman2022quantifying}.
Public universities are by-in-large not selective, so are caught in a relative decline when their selective and private peers are increasingly more selective, and better performing in terms of research and student outcomes.

While market forces have been effective in the successes of US higher education research, we should worry about the effects on education.
Enrolment in higher education has increased drastically since 1990, and public universities educate more than twice as many undergraduates in 2021 as their private counterparts.
%At the same time returns to higher education have increased drastically \citep{berman1998implications}.
Increasingly these students are being taught by lecturers, and not tenured faculty.
Lecturers are often employed on short-term or part-time contracts (adjunct), with limited job stability.
While we should worry for faculty, and their working conditions, there is credible evidence that relying on adjunct lecturers leads to worse student outcomes, relative to full-time lecturers and professors \citep{zhu2021limited}.
As tenure has increasingly become a private sector phenomenon, we should worry about the long-term effects on higher education, and the effect on the average student who attends a public university in the US.

%Additionally, put into further context of the trends in university competition \citep{urquiola2020markets}, and XYZ \citep{ehrenberg2012american}.

### 2023-07-13 text/sections/07-conclusion.tex
re-writing

This analysis investigates how the recent stagnation in state support for higher education has affected public universities, and their faculty composition.
This work contributes to the literature along two primary dimensions.
Firstly, by isolating changes in state funding on public universities via state funding shocks, the analysis provides a plausible mechanism for the observed negative effects on students outcomes that this stagnation has brought \citep{NBERw23736,NBERw27885}.
Secondly, this analysis considers individual professors as the unit of analysis using a dataset (IBHED) new in the economic literature.
These data allow for detailed analysis of thousand's of professors' salaries, and are a starting point for further analysis of the determinants of professors wages in the Illinois public university system.

Public universities have systematically substituted away from tenure-track and tenured professors, towards non-tenured lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, while suggestive information shows falls in hiring, which together implies that the changes in faculty composition are driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

While costs education have been rising in the US, public universities have also dealt with declining state funding.
It is natural to expect that such headwinds will lead to systematic change at public universities, changes that affect their faculty, and limit the universities' goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty composition, research at public universities, and the impact on higher education as a whole.

>>>>

This analysis investigates how the recent stagnation in state support for higher education has affected faculty, and their composition, at US public universities.
This work contributes to the literature along two primary dimensions.
Firstly, by isolating changes in state funding on public universities via state funding shocks, this work provides an explanation for the increased relaiance on lecturers and away from tenure-track and tenured professors at US public universities.
This approach used multiple methods to estimate the short- and medium-run effects of state funding cuts on faculty outcomes.
Secondly, this work asks investigates individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes in the Illinois university system.

I found that public universities have systematically substituted away from tenure-track and tenured professors, towards lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, while suggestive information shows falls in hiring, which together implies that the changes in faculty composition are driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

While costs education have been rising in the US, public universities have also dealt with declining state funding.
I show that these headwinds led to systematic change at public universities, changes that affect their faculty, likely limiting public universities in their goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty composition, research at public universities, and the impact on higher education as a whole.
